Do investors in clean energy ETFs herd?

### 2024-12-09 alt_energy_herding.qmd
cleaning up references

Argument: Investment interest on green ETFs has been explosive mainly by climate risks faced by the global economy and the actions of policy makers, goverments and organizations towards a sustainable future.

Derived from @decclesia2024: Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing ones in 2022, followed by the Cybersecurity and Artificial Intelligence (AI) ETFs.
The clean energy transition represents one of the largest multi-decade secular growth opportunities.

From @naqvi2022: After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products; all have escalated enormously.

@@

The sample consists of alternative energy equity ETFs (green ETFs) that are traded in the US markets (see @tbl-data in the Appendix).^[The data were sourced from \url{https://datastream.org/en-ca/}] The number of available alternative energy ETFs in our sample varied from 10 in the beginning of analysis to 30 at the most. The period of analysis runs from May 1^st^ of 2016 through 19^th^ June of 2024. The starting date was selected on the basis of the COP Paris agreement.
Daily logarithmic returns were computed from the closing prices of ETFs for a total of 2122 observations.

@@

Following the relevant literature we calculate the Cross Sectional Absolute Deviation  (CSAD) measure for each day in the following manner. We compute the difference of the ith ETF and market return where market return is proxied by the cross sectional average of returns for sample of our ETFs available for each day using @eq-csad:

$$
CSAD_t = \dfrac{1}{N}\sum_{i=1}^{N} |R_{i,t} - R_{m,t}|
$$ {#eq-csad}

```{r}
#| label: cross_sectional_deviation
#| fig.cap: "Cross Sectional Absolute Deviation (CSAD) for US Alternative Energy ETFs"
#| fig.height: 4
#| fig.width: 6
#| cache: false

At a later stage we estimate a non-linear regression as in @galariotis2015:

$$
CSAD_t = \gamma_0 + \gamma_1 R_{m,t} + \gamma_2 R_{m,t}^2 + \epsilon_t
$$ {#eq-herding}

The testing procedure of herding relies on the above @eq-herding. Rational asset pricing models predict a linear relationship between return dispersion and market returns under normal conditions, a relationship that is no longer valid in the presence of herding. Herding behaviour leads to an increasing or decreasing cross sectional dispersion with respect to market returns. In other words, herding is captured by a non-linear term in the standard pricing equation indicating a decreasing or an increasing returns’ dispersion.

@@

$$
CSAD_t = \gamma_0 + \gamma_1(1-D) R_{m,t} + \gamma_2D R_{m,t} + \gamma_3(1-D) R_{m,t}  + \gamma_4D R_{m,t}^2 \epsilon_t
$$ {#eq-asymmetry}

where $D$ is a dummy variable that takes the value of 1 when the market return is negative and 0 otherwise. Therefore, our exploration of asymmetric behavior of herding phenomenon is carried through the inspection of the statistical significance and the sign of the two estimated coefficients $\gamma_3$ vs $\gamma_4$ (up vs down markets).

@@

## General herding behaviour

The behaviour of the CSAD measure for US Alternative Energy ETFs is presented in @fig-csad.

@@

tribble(
  ~ "Quantile", ~ "gamma_0", ~ "gamma_1", ~ "gamma_2",
  "10%", "0.0016***", "0.2536***", " -1.3736",
  "25%", "0.0026***", "0.2461***", " -1.1056***",
  "50%", "0.0037***", "0.2648***", " -1.165***",
  "75%", "0.0048***", "0.3011***", " -1.1473***",
  "90%", "0.0064***", "0.2999***", " 0.2314"
) %>% 
  flextable_pdf() %>% 
  mk_par(i = 1, j = c(2, 3, 4), part = "header", value = as_paragraph(as_equation(headings))) %>% 
  mk_par(j = 1, part = "body", value = as_paragraph(as_equation(quantiles))) %>%
  add_footer_row(values = as_paragraph("Note: This table presents the estimation results of herding of US Alternative energy equity ETFs according to Equation 2 in various quantiles 10, 25, 50, 75 and 90% of the returns distribution.", " ", "*,**,*** denotes significance at 10%,5% and 1% respectively."
  ), colwidths = 4) %>% 
  hline_bottom(part = "footer", border = fp_border(color = "black", style = "solid", width = 1)) %>% 
  autofit(add_w = 0.2)

@@

## Rolling window analysis

@@

```{r}
#| label: fig-rolling_window
#| fig.cap: "Rolling window herding estimates"
#| cache: true
#| fig.width: 8
#| fig.height: 6

@@

## Probit analysis

The behaviour of participants in energy markets is closely related to the developments in the field of climate risks, carbon emissions  and environmentally friendly policies. There are a few studies that attempt to quantify the effects of uncertainty related to climate on the economy and financial markets \textcolor{red}{(see inter alia….)}. @bua2024 developed two climate risk related indexes namely transition and physical risk using a text-based approach in order to study the effect of these risks in financial markets.  It is expected that investors would prefer to hold assets that perform well in the face of increasing climate change risks, even if this entails accepting lower returns for such climate-hedging assets. Therefore, in the context of our study and following previous studies that study the determinants of herding behaviour [see @bouri2019; @demirer2018], we attempt to study the effect of climate-related uncertainty on the formation of herding behavior in the clean energy market. Given this, we define a dummy variable, which takes a value of 1 during periods of statistically significant herding (i.e., for days when the rolling t-statistic on $\gamma_2 < −1.96$) and zero otherwise, and then, we use a Probit model to relate this dummy to the two climate risk indexes developed by @bua2024. It should be noted that due to availability issues the probit analysis ends in December of 2023. The results from the Probit model are reported in @tbl-probit_analysis, where only the physical risk index significantly decreases the probability of herding. In other words, climate risks is good news for clean energy stocks or firms resulting in anti herding behaviour.

This implies that in the presence of higher physical risk with respect to the climate, clean energy ETFs become a more attractive investment option for investors that allocate their money to the various alternative energy investment products. As a result, the cross sectional dispersion of clean energy ETFs tends to increase.

@@

tribble(
  ~ "Variable", ~ "Coefficient",
  "Transition Risk", "-4.607***",
  "Physical Risk", "-1.318", 
  "Constant", " -1.506***",
  "LogL", "-484.7",
  "Observations with Dependent Variable (Dep) = 0", "1816",
  "Observations with Dependent Variable (Dep) = 1", "134"
) %>% 
  flextable_pdf() %>% 
  align(align = "left", j = 1, part = "all") %>%
  add_footer_lines(
    values = as_paragraph(
      "Notes: **,*** denotes statistically significant at 5% and 1%"
    )
  ) %>% 
  hline_bottom(part = "footer", border = fp_border(color = "black", style = "solid", width = 1)) %>% 
  autofit(add_w = 1.5)

@@

Furthermore, suppose we get the median values of these two series. then we define values above median as high and below median as low. We define a dummy =1 if values>median and 0 otherwise and in this way we will have PRI values that are bigger than median and 0 otherwise. We also define a dummy which is lower than median, i.e., 1 and 0 otherwise, then we use these high PRI and high TRI in one probit regression and low TRI and low PRI in another. Results are presented in the following table. We observe higher uncertainty that stems from physical or transition risk causes anti-herding which is in line with the logic we discussed earlier.

tribble(
  ~" ", ~ "High", ~ "Low",
  "Physical Risk", "-6.736*", "-6.118",
  "Transition Risk", "-1.798", "-2.581",
) %>% 
  flextable_pdf() %>% 
  align(align = "left", j = 1, part = "all") %>%
  add_footer_lines(
    values = as_paragraph(
      "Notes: *, denotes statistically significant at 10%"
    )
  ) %>%
  hline_bottom(part = "footer", border = fp_border(color = "black", style = "solid", width = 1)) %>% 
  autofit(add_w = 1.4)

@@

```{r}
#| label: tbl-data
#| tbl-cap: "List of used alternative energy equity ETFs"
#| cache: false

tribble(
  ~"ETF",
  "ALPS CLEAN ENERGY ETF",
  "BLUE HORIZON BNE ETF",
  "SPDR S&P KENSHO CLEAN POWER ETF",
  "GLOBAL X CLEANTECH ETF",
  "PROSHARES S&P KENSHO CLEANTECH ETF",
  "INVESCO MSCI SUSTAINABLE FUTURE ETF",
  "FIRST TRUST GLOBAL WIND ENERGY ETF",
  "FIDELITY CLEAN ENERGY ETF",
  "GLDS.BLOOMBERG CN. EN. EQ.ETF",
  "FST.NQ.CN.EDGE SMRT.GRID INFRA IDX ETF",
  "DEFIANCE NEXT GEN H2 ETF",
  "DIREXION HYDROGEN ETF",
  "GLOBAL X HYDROGEN ETF",
  "ISHARES GLOBAL CLEAN EN. ETF",
  "BLACKR.WLD.EXUS CRBN TSTN.READINESS",
  "NUB.CBN.TSTN.& INFRA",
  "TCW TRANSFORM SYSTEMS ETF",
  "VANECK URANIUM AND NUCLEAR ENERGY",
  "NUVEEN GLOBAL NET ZERO TRANSITION ETF",
  "SPDR MSCI USA CIM. PA. ALIGNED ETF", 
  "INVESCO GLOBAL CLEAN ENERGY ETF", 
  "FST.NQ.CN.EDGE GREY.ETF",
  "GLOBAL X SOLAR ETF", 
  "GLOBAL X RENEWABLE ENERGY PRODUCERS",
  "TRUESHARES EAG.GLB. RENWEN.ETF",
  "VANECK LOW CARBON ENERGY ETF",
  "SMARTETFS SUST.EN. II ETF",
  "INVESCO SOLAR ETF", 
  "VIRTUS DUFF & PHELPS CLEAN ENERGY ETF",
  "GLOBAL X WIND ENERGY ETF"
) %>% 
  flextable_pdf() %>% 
  align(align = "left", j = 1,  part = "all") %>%
  autofit(add_w = 3)

@@

>>>>

Investment interest on green ETFs has been explosive mainly by climate risks faced by the global economy and the actions of policy makers, governments and organizations towards a sustainable future.  Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing ones in 2022, followed by the Cybersecurity and Artificial Intelligence (AI) ETFs [@decclesia2024]. The clean energy transition represents one of the largest multi-decade secular growth opportunities.  After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products; all have escalated enormously [@naqvi2022].

@@

The sample consists of alternative energy equity ETFs (green ETFs) that are traded in the US markets (see @tbl-data in the Appendix).^[The data were sourced from \url{https://datastream.org/en-ca/}] The number of available alternative energy ETFs in our sample varied from 10 in the beginning of analysis to 30 at the most. The period of analysis runs from May 1^st^ of 2016 through 19^th^ June of 2024. The starting date was selected on the basis of the UN Climate Change Conference (COP) Paris agreement.
Daily logarithmic returns were computed from the closing prices of ETFs for a total of 2122 observations.

@@

Following the relevant literature [@christie1995; and @chang2000], we compute dispersion of the $ith$ ETF from the market return. This is known as the Cross Sectional Absolute Deviation  ($CSAD_t$) measure. Empirically the $CSAD_t$ is defined in the following manner:

$$
CSAD_t = \dfrac{1}{N}\sum_{i=1}^{N} |R_{i,t} - R_{m,t}|,
$$ {#eq-csad}

where $R_{i,t}$ is the return and $R_{m,t}$ is the cross sectional average of returns for the sample of ETFs available for each day. The return dispersion measures the directional similarity of ETF returns to the market return. This return similarity forms the basis for the herding behaviour tests. The CSAD measure for US Alternative Energy ETFs is presented in @fig-csad. Following @galariotis2015 we estimate @eq-herding:

$$
CSAD_t = \gamma_0 + \gamma_1 R_{m,t} + \gamma_2 R_{m,t}^2 + \epsilon_t,
$$ {#eq-herding}

where $\gamma_0$ is the intercept, $\gamma_1$ is the coefficient of the linear term, $\gamma_2$ is the coefficient of the quadratic term or the herding behaviour term, and $\epsilon_t$ is the error term. The coefficient $\gamma_2 <0$ when herding is present, and $\gamma_2 >0$  when anti-herding is present. To ensure the robustness of the estimate, we estimate $CSAD_{t}$ with Newey-West standard errors [See @newey1987].

@@

$$
CSAD_t = \gamma_0 + \gamma_1(1-D) R_{m,t} + \gamma_2D R_{m,t} + \gamma_3(1-D) R_{m,t}^2  + \gamma_4D R_{m,t}^2 + \epsilon_t,
$$ {#eq-asymmetry}

where $D$ is a dummy variable that takes the value of 1 when the market return is negative and 0 otherwise. Therefore, our exploration of asymmetric behaviour of herding phenomenon is carried through the inspection of the statistical significance and the sign of the two estimated coefficients $\gamma_3$ versus $\gamma_4$ (up versus down markets).

@@

## Herding behaviour
Rational asset pricing models [for example, @black1972] predict a linear relationship between return dispersion and market returns under normal conditions, a relationship that is no longer valid in the presence of herding. Herding behaviour leads to an increasing or decreasing cross sectional dispersion with respect to market returns. In other words, herding is captured by a non-linear term in the standard pricing equation indicating a decreasing or an increasing returns’ dispersion.

@@

tribble(
  ~ "Quantile", ~ "gamma_0", ~ "gamma_1", ~ "gamma_2",
  "10%", "0.0016***", "0.2536***", " -1.3736",
  "25%", "0.0026***", "0.2461***", " -1.1056***",
  "50%", "0.0037***", "0.2648***", " -1.165***",
  "75%", "0.0048***", "0.3011***", " -1.1473***",
  "90%", "0.0064***", "0.2999***", " 0.2314"
) %>% 
  flextable_pdf() %>% 
  mk_par(i = 1, j = c(2, 3, 4), part = "header", value = as_paragraph(as_equation(headings))) %>% 
  mk_par(j = 1, part = "body", value = as_paragraph(as_equation(quantiles))) %>%
  add_footer_row(values = as_paragraph("Note: This table presents the estimation results of herding of US Alternative energy equity ETFs according to Equation 2 in various quantiles 10, 25, 50, 75 and 90% of the returns distribution.", " ", "*,**,*** denotes significance at 10%,5% and 1% respectively."
  ), colwidths = 4) %>% 
  hline_bottom(part = "footer", border = fp_border(color = "black", style = "solid", width = 1)) %>% 
  autofit(add_w = 0.3)

@@

## Herding behaviour over-time

@@

```{r}
#| label: fig-rolling_window
#| fig.cap: "Rolling window herding estimates. Note: The red perforated lines indicates the 95% confidence interval."
#| cache: true
#| fig.width: 8
#| fig.height: 6

@@

## Climate-related uncertainty and herding behaviour

The behaviour of participants in energy markets is closely related to the developments in the field of climate risks, carbon emissions  and environmentally friendly policies. There are a few studies that attempt to quantify the effects of uncertainty related to climate on the economy and financial markets \textcolor{red}{(see inter alia….)}. @bua2024 developed two climate risk related indexes namely transition and physical risk using a text-based approach in order to study the effect of these risks in financial markets.  It is expected that investors would prefer to hold assets that perform well in the face of increasing climate change risks, even if this entails accepting lower returns for such climate-hedging assets. Therefore, in the context of our study and following previous studies that study the determinants of herding behaviour [see @bouri2019; @demirer2018], we attempt to study the effect of climate-related uncertainty on the formation of herding behavior in the clean energy market.

We use a probit model to relate herding to the two climate risk indexes developed by @bua2024 in the following manner:

$$
Pr(D^{herd} = 1 |\lambda_0 +  \lambda_1TRI  + \lambda_2PRI) = \lambda_0 +  \lambda_1TRI  + \lambda_2PRI,
$$ {#eq-climate}

where $D^{herd}$ takes a value of 1 during periods of statistically significant herding (i.e., for days when the rolling t-statistic on $\gamma_2 < −1.96$ in @fig-rolling_window) and zero otherwise. $TRI$ is the transitional risk index and $PRI$ is the physical risk index.

The results from the Probit model are reported in @tbl-probit_analysis, where only the physical risk index significantly decreases the probability of herding.^[It should be noted that due to availability issues the probit analysis ends in December of 2023.] In other words, climate risks is good news for clean energy stocks or firms resulting in anti-herding behaviour. This implies that in the presence of higher physical risk with respect to the climate, clean energy ETFs become a more attractive investment option for investors that allocate their money to the various alternative energy investment products. As a result, the cross sectional dispersion of clean energy ETFs tends to increase.

lambda <- c(
"\\lambda_0",
"\\lambda_1",
"\\lambda_2"
)

tribble(
  ~ "Variable", ~ "Coefficient",
  "lambda_0", " -1.506***",
  "lambda_1", "-4.607***",
  "lambda_2", "-1.318", 
  "Log Likelihood", "-484.7",
  "Observations with Dependent Variable (Dep) = 0", "1816",
  "Observations with Dependent Variable (Dep) = 1", "134"
) %>% 
  flextable_pdf() %>% 
  align(align = "left", j = 1, part = "all") %>%
  mk_par(j = 1, i = c(1, 2, 3), part = "body", value = as_paragraph(as_equation(lambda))) %>% 
  add_footer_lines(
    values = as_paragraph(
      "Notes: **,*** denotes statistically significant at 5% and 1%"
    )
  ) %>% 
  hline(part = "body", i = 3, border = fp_border(color = "black", style = "solid", width = 1)) %>% 
  hline_bottom(part = "footer", border = fp_border(color = "black", style = "solid", width = 1)) %>% 
  autofit(add_w = 1.5)

Furthermore, we develop two additional models to study the effect of climate risks on herding behaviour. We split the sample into two groups based on the median value of the TRI and PRI. We then estimate the following two models:

$$
Pr(D^{herd} = 1 |\lambda_0 +\lambda_1D^{TRI}_{high}TRI + \lambda_2D^{PRI}_{high}PRI) = \lambda_0 +\lambda_1D^{TRI}_{high}TRI + \lambda_2D^{PRI}_{high}PRI, and
$$ {#eq-climate_high}

$$
Pr(D^{herd} = 1 |\lambda_0 +\lambda_1D^{TRI}_{low}TRI + \lambda_2D^{PRI}_{low}PRI) = \lambda_0 +\lambda_1D^{TRI}_{low}TRI + \lambda_2D^{PRI}_{low}PRI, 
$$ {#eq-climate_low}

where $D^{herd}$ is the same as in @eq-climate. $D^{TRI}_{high}$ and $D^{PRI}_{high}$ are dummy variables that take a value of 1 if the value of the TRI and PRI are above the median and zero otherwise. Similarly, $D^{TRI}_{low}$ and $D^{PRI}_{low}$ are dummy variables that take a value of 1 if the value of the TRI and PRI are below the median and zero otherwise.

@@

lambda <- c(
"\\lambda_1",
"\\lambda_2"
)

tribble(
  ~" ", ~ "High", ~ "Low",
  "Physical Risk", "-6.736*", "-6.118",
  "Transition Risk", "-1.798", "-2.581",
) %>% 
  flextable_pdf() %>% 
  align(align = "left", j = 1, part = "all") %>%
  add_footer_lines(
    values = as_paragraph(
      "Notes: *, denotes statistically significant at 10%"
    )
  ) %>%
  mk_par(j = 1, i = c(1, 2), part = "body", value = as_paragraph(as_equation(lambda))) %>% 
  hline_bottom(part = "footer", border = fp_border(color = "black", style = "solid", width = 1)) %>% 
  autofit(add_w = 1.4)

@@

We use these high PRI and high TRI in one probit regression and low TRI and low PRI in another. Results are presented @tbl-probit_analysis2. We observe higher uncertainty that stems from physical or transition risk causes anti-herding which is in line with the logic we discussed earlier.

@@

## Cross-sectional average deviations

@@

## ETFs used in the study

```{r}
#| label: tbl-data
#| tbl-cap: "List of green ETFs used in the study"
#| cache: false

tribble(
  ~"ETF",
  "ALPS CLEAN ENERGY ETF",
  "BLUE HORIZON BNE ETF",
  "SPDR S&P KENSHO CLEAN POWER ETF",
  "GLOBAL X CLEANTECH ETF",
  "PROSHARES S&P KENSHO CLEANTECH ETF",
  "INVESCO MSCI SUSTAINABLE FUTURE ETF",
  "FIRST TRUST GLOBAL WIND ENERGY ETF",
  "FIDELITY CLEAN ENERGY ETF",
  "GLDS.BLOOMBERG CN. EN. EQ.ETF",
  "FST.NQ.CN.EDGE SMRT.GRID INFRA IDX ETF",
  "DEFIANCE NEXT GEN H2 ETF",
  "DIREXION HYDROGEN ETF",
  "GLOBAL X HYDROGEN ETF",
  "ISHARES GLOBAL CLEAN EN. ETF",
  "BLACKR.WLD.EXUS CRBN TSTN.READINESS",
  "NUB.CBN.TSTN.& INFRA",
  "TCW TRANSFORM SYSTEMS ETF",
  "VANECK URANIUM AND NUCLEAR ENERGY",
  "NUVEEN GLOBAL NET ZERO TRANSITION ETF",
  "SPDR MSCI USA CIM. PA. ALIGNED ETF", 
  "INVESCO GLOBAL CLEAN ENERGY ETF", 
  "FST.NQ.CN.EDGE GREY.ETF",
  "GLOBAL X SOLAR ETF", 
  "GLOBAL X RENEWABLE ENERGY PRODUCERS",
  "TRUESHARES EAG.GLB. RENWEN.ETF",
  "VANECK LOW CARBON ENERGY ETF",
  "SMARTETFS SUST.EN. II ETF",
  "INVESCO SOLAR ETF", 
  "VIRTUS DUFF & PHELPS CLEAN ENERGY ETF",
  "GLOBAL X WIND ENERGY ETF"
) %>% 
  flextable_pdf() %>% 
  align(align = "left", j = 1,  part = "all") %>%
  add_footer_row(values = as_paragraph("Note: Details on this funds can be found on Yahoo Finance."), colwidths = 1) %>% 
  autofit(add_w = 3)
